---
title: "Tissue sampling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tissue sampling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rRACES)
library(dplyr)
library(ggplot2)
```

Once one has familiarised on how a tumour evolution simulation can be
programmed using rRACES, the next step is to augment the simulation with
sampling of tumour cells. This mimics a realistic experimental design
where we gather tumour sequencing data.

This vignette introduces sampling using different type of models;
starting from simpler up to more complex simulation scenarios we
consider:

-   *multi-region sampling*: where at every time point multiple
    spatially-separated samples are collected;

-   *longitudinal sampling*: where the sampling is repeated at multiple
    time-points.

## Custom multi-region sampling

We consider a simple monoclonal model, without epigenotypes.

```{r, eval=TRUE}
# Monoclonal model, no epigenotypes
sim <- new(Simulation, "Monoclonal")

sim$add_genotype(name = "A", growth_rates = 0.1, death_rates = 0.01)

sim$place_cell("A", 500, 500)
sim$run_up_to_size("A", 60000)
```

```{r,  fig.height=4, fig.width=3}
current <- plot_tissue(sim)
current
```

A sample is defined by a name and a bounding box, which has a $(x,y)$
coordinate for the bottom left point, and for the top right point.

For this simulation, we define two samples with names `"S_1_2"` and
`"S_1_2"`.

```{r,  fig.height=4, fig.width=3}
# We collect a squared box of (bbox_width x bbox_width) cells
bbox_width <- 50

# Box A1
bbox1_p <- c(400, 400)
bbox1_q <- bbox1_p + bbox_width

# Box B1
bbox2_p <- c(500, 500)
bbox2_q <- bbox2_p + bbox_width

# View the boxes
current +
  geom_rect(xmin = bbox1_p[1], xmax = bbox1_q[2],
            ymin = bbox1_p[1], ymax = bbox1_q[2],
            fill = NA, color = "black") +
  geom_rect(xmin = bbox2_p[1], xmax = bbox2_q[2],
            ymin = bbox2_p[1], ymax = bbox2_q[2],
            fill = NA, color = "black")

# Sampling
sim$sample_cells("S_1_1", bottom_left = bbox1_p, top_right = bbox1_q)
sim$sample_cells("S_1_2", bottom_left = bbox2_p, top_right = bbox2_q)
```

> *Note:* Sampling removes cells from the tissue, as if the tissue was
> surgically resected. Therefore, cells that are mapped to the bounding
> box after application of `sample_cells` are no longer part of the
> simulation.

A new call to `plot_tissue` will show the box where the cells have been
removed to be white.

```{r,  fig.height=4, fig.width=3}
plot_tissue(sim)
```

This is also reflected by `get_cells`, which now will not find any
tumour cell in the sampled region.

```{r}
# This should be empty
sim$get_cells(c(400, 400), c(400 + bbox_width, 400 + bbox_width)) %>% head
```

### Cell division tree for sampled cells

Every sampled cell is linked, at the evolutionary level,
to the other cells that originate from
the same initial cell. It helps to visualise the evolutionary information on the cells that we have sampled as a forest of trees (if one seeded
multiple initial cells). The forest is an object of the S4 class
`SamplesForest`.

```{r}
forest <- sim$get_samples_forest()

forest
```

The forest has methods to obtain the nodes of the sampled cells.

```{r}
forest$get_nodes() %>% head
```

The leaves of the forest are  sampled cells, while the
internal nodes are their ancestors. The field `sample`
is not available for internal nodes, and reports the
sample name otherwise.

```{r}
# The leaves in the forest represent sampled cells
forest$get_nodes() %>%
  filter(!is.na(.data$sample)) %>%
  head
```

The roots of the forest having no ancestors.

```{r}
# If it is one cell, than the forest is a tree
forest$get_nodes() %>%
  filter(is.na(.data$ancestor))
```

We can also query the forest about the samples used to build it.

```{r}
forest$get_samples_info()
```

We can visualise the forest.
This plot reports the cells and, on the y-axis, their time of birth.
```{r,  fig.height=6, fig.width=11}
plot_forest(forest)
```
The plot shows also samples annotations and species but, for a large number of cells, it might be complicated to view the full tree, unless a very large canvas is used. For this reaason, it is possible to subset the tree.

```{r,  fig.height=6, fig.width=11}
# Extract the subforest linked to sample
S_1_1_forest <- forest$get_subforest_for("S_1_1")

plot_forest(S_1_1_forest)
```

In general, these plots can be annotated with extra information, such as the sampling times, and the MRCAs of each sample in the tree.

```{r,  fig.height=6, fig.width=11}
# Full plot
plot_forest(forest) %>%
  annotate_forest(forest)
# S_1_1 plot
plot_forest(S_1_1_forest) %>%
  annotate_forest(S_1_1_forest)
```

## Randomised multi-region samples


```{r}
sim <- new(Simulation, "Randomised")

sim$add_genotype(name = "A", growth_rates = 0.1, death_rates = 0.01)

sim$place_cell("A", 500, 500)
sim$run_up_to_size("A", 60000)
```

We include a new clone and let it grow. This new clone has much higher
growth rates than its ancestor.

```{r,  fig.height=4, fig.width=3}
# Add a new clone
sim$add_genotype(name = "B", growth_rates = 1, death_rates = 0.01)
sim$mutate_progeny(sim$choose_cell_in("A"), "B")

sim$run_up_to_size("B", 10000)

current <- plot_tissue(sim)
current
```

Since clone start has been randomised by `choose_cell_in`, we have no exact idea of where to sample to obtain for example, $100$ of its cells. We can look visually at the simulation, but this is slow.

rRACES provides a `bbox_sampler` function to sample bounding boxes that
contain a desired number of cells. The function takes in input:

- a bounding box size;
- the number $n$ of cells to sample for a species of interest.

`bbox_sampler` will attempt a fixed number of times to sample the box,  starting from positions occupied by the species of interest. If  a box that contains at least $n$ cells is not found within a number of attempts, then the one with the largest number of samples is returned.

This allows to program sampling without having a clear idea of the tissue conformation.

```{r,  fig.height=4, fig.width=3}
# A bounding box 50x50 with at least 100 cells of species B
ncells <- 100
n_w <- n_h <- 50

# Sampling ncells with random box sampling of boxes of size n_w x n_h
bbox <- bbox_sampler(sim, "B", ncells, n_w, n_h)

current +
  geom_rect(xmin = bbox$p[1], xmax = bbox$q[1],
            ymin = bbox$p[2], ymax = bbox$q[2],
            fill = NA, color = "black")
```

```{r}
sim$sample_cells("S_2_1", bbox$p, bbox$q)
```

Something similar with species `A`.

```{r,  fig.height=4, fig.width=3}
bbox <- bbox_sampler(sim, "A", ncells, n_w, n_h)

current +
  geom_rect(xmin = bbox$p[1], xmax = bbox$q[1],
            ymin = bbox$p[2], ymax = bbox$q[2],
            fill = NA, color = "black")
```

```{r}
sim$sample_cells("S_2_2", bbox$p, bbox$q)
```

The two samples have been extracted.

```{r,  fig.height=4, fig.width=3}
plot_tissue(sim)
```

Cell division tree.

```{r,  fig.height=6, fig.width=11}
forest <- sim$get_samples_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

## Two populations with epigenetic state

We are now ready to simulate a model with epigenetic
switches and subclonal expansions.

```{r,  fig.height=4, fig.width=3}
sim <- new(Simulation, "Two Populations")

sim$death_activation_level <- 20

# First clone
sim$add_genotype(name = "A",
                 epigenetic_rates = c("+-" = 0.01, "-+" = 0.01),
                 growth_rates = c("+" = 0.1, "-" = 0.08),
                 death_rates = c("+" = 0.1, "-" = 0.01))

sim$place_cell("A+", 500, 500)
sim$run_up_to_size("A+", 1000)
plot_tissue(sim, num_of_bins = 500)
```

We sample before introducing a new clone.

```{r,  fig.height=4, fig.width=3}
bbox_width <- 10

sim$sample_cells("A1",
                 bottom_left = c(480, 480),
                 top_right = c(480 + bbox_width, 480 + bbox_width))

sim$sample_cells("B1",
                 bottom_left = c(500, 500),
                 top_right = c(500 + bbox_width, 500 + bbox_width))

plot_tissue(sim, num_of_bins = 500)
```

```{r,  fig.height=4, fig.width=3}
# Let it grow a bit more
sim$run_up_to_time(sim$get_clock() + 15)

plot_tissue(sim, num_of_bins = 500)
```

Add a new subclone.

```{r,  fig.height=4, fig.width=3}
cell <- sim$choose_cell_in("A")

sim$add_genotype(name = "B",
                 epigenetic_rates = c("+-" = 0.05, "-+" = 0.1),
                 growth_rates = c("+" = 0.5, "-" = 0.3),
                 death_rates = c("+" = 0.05, "-" = 0.05))

sim$mutate_progeny(cell, "B")

# let it grow more time units
sim$run_up_to_time(sim$get_clock() + 25)

plot_tissue(sim, num_of_bins = 500)
```

Sample again and plot the tissue

```{r,  fig.height=4, fig.width=3}

sim$sample_cells("A2", bottom_left = c(250, 500),
                 top_right = c(250 + bbox_width, 500 + bbox_width))

sim$sample_cells("B2", bottom_left = c(500, 165),
                 top_right = c(500 + bbox_width, 165 + bbox_width))

plot_tissue(sim, num_of_bins = 500)
plot_muller(sim)
```

Now we show the cell division tree, which starts being rather
complicated

```{r,  fig.height=12, fig.width=17}
forest <- sim$get_samples_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

## More complex example

Some features allow for a smarter sampling and more advanced
visualisation.

We build a tumour with three populations, branching.

```{r, fig.height=4, fig.width=3}
sim <- new(Simulation, "Three Populations")

sim$death_activation_level <- 20

sim$update_tissue(250, 250)

sim$add_genotype(name = "A", growth_rates = 0.08, death_rates = 0.01)
sim$place_cell("A", 125, 125)
sim$run_up_to_size("A", 300)

plot_tissue(sim)
```

We use an automatic function `bbox_sampler` to determine bounding boxes
so to collect a dedired number of cells of interest.

```{r,  fig.height=4, fig.width=3}
ncells <- 20
n_w <- n_h <- 7

# Sampling ncells with random box sampling of boxes of size n_w x n_h
bbox <- bbox_sampler(sim, "A", ncells, n_w, n_h)
sim$sample_cells("A1", bbox$p, bbox$q)

bbox2 <- bbox_sampler(sim, "A", ncells, n_w, n_h)
sim$sample_cells("B1", bbox2$p, bbox2$q)

plot_tissue(sim)
```

```{r,  fig.height=4, fig.width=3, eval=FALSE}
plot_muller(sim)
```

Second clone starts.

```{r,  fig.height=4, fig.width=3}
# New subclone
sim$add_genotype(name = "B", growth_rates = 0.13, death_rates = 0.01)
sim$mutate_progeny(sim$choose_cell_in("A"), "B")
sim$run_up_to_size("B", 200)

plot_tissue(sim)
plot_muller(sim)
```

Random sampling of cells from both clone `A` and `B`.

```{r,  fig.height=4, fig.width=3}
# Sampling with random box sampling
bbox <- bbox_sampler(sim, "A", ncells, n_w, n_h)
sim$sample_cells("A2", bbox$p, bbox$q)

bbox <- bbox_sampler(sim, "B", ncells, n_w, n_h)
sim$sample_cells("B2", bbox$p, bbox$q)

plot_tissue(sim)
```

Finally, a third clone branching from `A`.

```{r,  fig.height=4, fig.width=3}
# New subclone
sim$add_genotype(name = "C", growth_rates = 0.6, death_rates = 0.01)
sim$mutate_progeny(sim$choose_cell_in("A"), "C")

sim$run_up_to_size("C", 500)
plot_tissue(sim)
plot_muller(sim)
```

With sampling of clones `B` and `C`.

```{r,  fig.height=4, fig.width=3}
# Sampling with random box sampling
bbox <- bbox_sampler(sim, "B", ncells, n_w, n_h)
sim$sample_cells("A3", bbox$p, bbox$q)

bbox <- bbox_sampler(sim, "C", ncells, n_w, n_h)
sim$sample_cells("B3", bbox$p, bbox$q)

plot_tissue(sim, num_of_bins = 100)
plot_muller(sim)
```

Finally, the tree plot where we annotate samples and MRCAs of each
sample.

```{r,  fig.height=12, fig.width=17}
# Tree data
forest <- sim$get_samples_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

The horizontal line denotes sampling times, and circles the MRCAs of interest.

