---
title: "Growth Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Growth Models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

rRACES/RACES supports two growth models: the "border"-growth model and the homogeneous-growth
model. 

The former exclusively admits duplication of cells that have space to duplicate 
(i.e., cells on the cancer border or near by death cells). The latter allows cell
duplications everywhere in the tumor.


The homogeneous-growth model is the default one. However, users can switch the 
growth model by assigning the `Simulation`'s `duplicate_internal_cells` Boolean field.
By setting it to `TRUE`, the homogeneous-growth model is used. If, instead, it is 
set to `FALSE`, the simulation evolves according with the "border"-growth model.

In the remaining part of this article we clarify the differences between the 
supported growth models by showing that they produce different evolutions 
of the very same cancer model.

### Homogeneous Growth

First of all we need to create a new `Simulation` object and 
enable duplication of internal cells.

```{r}
library(rRACES)

sim <- new(Simulation, "Homogeneous Growth")

# Set the homogeneous growth model
sim$duplicate_internal_cells <- TRUE

# Set the death activation level to avoid drift
sim$death_activation_level <- 50
```

Add a mutant `A` with epigenetic state, let the 
simulation evolve until there are 1000 cells of species `A+`, 
take two samples, and let the simulation evolve again for 15 time units.

```{r}
# Add a mutant
sim$add_mutant(name = "A",
                 epigenetic_rates = c("+-" = 0.01, "-+" = 0.01),
                 growth_rates = c("+" = 0.1, "-" = 0.08),
                 death_rates = c("+" = 0.1, "-" = 0.01))

sim$place_cell("A+", 500, 500)

# Let the simulation evolve until "A+" consists of 1000 cells
sim$run_up_to_size("A+", 1000)

bbox_width <- 15

# Takes two samples
sim$sample_cells("S_1_1",
                 bottom_left = c(480, 480),
                 top_right = c(480 + bbox_width, 480 + bbox_width))

sim$sample_cells("S_1_2",
                 bottom_left = c(500, 500),
                 top_right = c(500 + bbox_width, 500 + bbox_width))

# Let the simulation evolve again for 15 time units
sim$run_up_to_time(sim$get_clock() + 15)
```

Add a new mutant "B", let one of the cells in "A" 
generate a cell in "B", let the simulation evolve until there 
are 5000 cells in "B+", and take again two samples.

```{r}
sim$add_mutant(name = "B",
                 epigenetic_rates = c("+-" = 0.05, "-+" = 0.1),
                 growth_rates = c("+" = 0.8, "-" = 0.3),
                 death_rates = c("+" = 0.05, "-" = 0.05))

sim$mutate_progeny(sim$choose_cell_in("A"), "B")

sim$run_up_to_size("B+", 5000)

ncells <- 0.9 * bbox_width * bbox_width

bbox <- sim$search_sample("B", ncells, bbox_width, bbox_width)
sim$sample_cells("S_2_1", bbox$lower_corner, bbox$upper_corner)

bbox <- sim$search_sample("A", ncells, bbox_width, bbox_width)
sim$sample_cells("S_2_2", bbox$lower_corner, bbox$upper_corner)
```

Let us have a look at the simulated tissue and plot the simulation Muller plot.

```{r,  fig.height=4, fig.width=3}
plot_tissue(sim, num_of_bins = 500)
plot_muller(sim)
```

Finally, let us build the ancestor forest of the samples.

```{r, fig.width=14}
library(dplyr)
library(ggplot2)

forest <- sim$get_samples_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

We use a special `highlight` parameter to add the edges connecting cells in each sample.
```{r, fig.width=14}
plot_forest(forest, highlight ='S_1_1') %>%
  annotate_forest(forest)

plot_forest(forest, highlight ='S_1_2') %>%
  annotate_forest(forest)

plot_forest(forest, highlight ='S_2_1') %>%
  annotate_forest(forest)

plot_forest(forest, highlight ='S_2_2') %>%
  annotate_forest(forest)
```




### "Border" Growth

We now build a simulation and switch to the "border" growth model.

```{r}
sim <- new(Simulation, "Border Growth")

# Set the "border" growth model
sim$duplicate_internal_cells <- FALSE

# Set the death activation level to avoid drift
sim$death_activation_level <- 50
```

Let us repeat what we did in the homogeneous growth model example.

```{r}
# Add a mutant
sim$add_mutant(name = "A",
                 epigenetic_rates = c("+-" = 0.01, "-+" = 0.01),
                 growth_rates = c("+" = 0.1, "-" = 0.08),
                 death_rates = c("+" = 0.1, "-" = 0.01))

sim$place_cell("A+", 500, 500)

# Let the simulation evolve until "A+" consists of 1000 cells
sim$run_up_to_size("A+", 1000)

bbox_width <- 15

# Takes two samples
sim$sample_cells("S_1_1",
                 bottom_left = c(480, 480),
                 top_right = c(480 + bbox_width, 480 + bbox_width))

sim$sample_cells("S_1_2",
                 bottom_left = c(500, 500),
                 top_right = c(500 + bbox_width, 500 + bbox_width))

# Let the simulation evolve again for 15 time units
sim$run_up_to_time(sim$get_clock() + 15)

# Add a new mutant
sim$add_mutant(name = "B",
                 epigenetic_rates = c("+-" = 0.05, "-+" = 0.1),
                 growth_rates = c("+" = 0.8, "-" = 0.3),
                 death_rates = c("+" = 0.05, "-" = 0.05))


# Let one of the "A" cells generate a cell in "B"
sim$mutate_progeny(sim$choose_cell_in("A"), "B")

# Let the simulation evolve until "B+" consists of 5000 cells
sim$run_up_to_size("B+", 5000)

ncells <- 0.8 * bbox_width * bbox_width

bbox <- sim$search_sample("B", ncells, bbox_width, bbox_width)
sim$sample_cells("S_2_1", bbox$lower_corner, bbox$upper_corner)

bbox <- sim$search_sample("A", ncells, bbox_width, bbox_width)
sim$sample_cells("S_2_2", bbox$lower_corner, bbox$upper_corner)
```

Let us have a look at the simulated tissue and plot the simulation Muller plot.

```{r,  fig.height=4, fig.width=3}
plot_tissue(sim, num_of_bins = 500)
plot_muller(sim)
```

Once more, let us build the ancestor forest of the samples.

```{r, fig.width=14}
forest <- sim$get_samples_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

We use a special `highlight` parameter to add the edges connecting cells in each sample.
```{r, fig.width=14}
plot_forest(forest, highlight ='S_1_1') %>%
  annotate_forest(forest)

plot_forest(forest, highlight ='S_1_2') %>%
  annotate_forest(forest)

plot_forest(forest, highlight ='S_2_1') %>%
  annotate_forest(forest)

plot_forest(forest, highlight ='S_2_2') %>%
  annotate_forest(forest)
```

It is easy to see the differences in the analoguous plots of the above examples.

First of all, the cancer growth is slower when subject to the "border"-growth model 
than when the homogeneous-growth model is used. This is due to the fact that 
the internal cells cannot duplicate in the former growth model, thus, the number of 
cells active for duplication is always greater in it, than in the homogeneous-growth 
model.

Moreover, spacial closeness and closeness in the samples ancestor forest 
are strictly related in the "border" growth model, whereas they appear to be loosely 
related in the homogeneous growth model. 
This features can be easily spotted in the forest plots. 
