---
title: "rRACES"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rRACES}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Using `rRACES`

rRACES is at early stage of development and misses many features of 
RACES. At the moment, it supports:
- creating a spacial simulation
- adding species to the simulation
- adding cells to the spacial simulation
- running a simulation up to a specified simulated time

Further classes and methods will be available soon.

## A Small Example

```{r, eval=FALSE}
require(rRACES)

# Create a new simulation and set the simulation data directory
# to be "output_dir". A 1000x1000-cells tissue is automatically
# added to the simulation
sim <- new(Simulation, "output_dir")

# Add two species, "A" and "B", having epigenetic states
sim$add_species(name = "A",
                epigenetic_rates = c("+-" = 0.01, "-+" = 0.01),
                growth_rates = c("+" = 0.2, "-" = 0.08),
                death_rates = c("+" = 0.1, "-" = 0.01))

sim$add_species(name = "B",
                epigenetic_rates = c("+-" = 0.01, "-+" = 0.01),
                growth_rates = c("+" = 0.1, "-" = 0.3),
                death_rates = c("+" = 0.05, "-" = 0.1))


# Program a timed transition from species "A" to
# species "B" at time 60
sim$add_timed_mutation("A", "B", 60)

# Get the simulation species with epigenetic state names
sim$get_species_names()

# Add one cell to the simulated tissue
sim$add_cell(genotype_name = "A+", x = 500, y = 500)

# Run the simulation up to time 110
sim$run_up_to(110)

# Get simulation time (you will get 110)
sim$get_clock()

# Get the cells in the tissue at current simulation time
sim$get_cells()

# Get the cells in the tissue rectangular sample having [500,500]
# and [505,505] as lower and upper corners, respectively
sim$get_cells(c(500, 500), c(505, 505))

# Get the cells in the tissue having epigenetic state "-"
sim$get_cells(c("A", "B"), c("-"))

# Get the cells in the tissue having epigenetic state "-" and, at
# the same time, belonging to rectangular sample bounded by [500,500]
# and [505,505] as lower and upper corners, respectively
sim$get_cells(c(500, 500), c(505, 505), c("A", "B"), c("-"))

# Report the number of cells per species at the current simulation time
sim$get_counts()

# Get the simulation directory
sim$get_directory()

# Get the tissue name
sim$get_tissue_name()

# Get the tissue size
sim$get_tissue_size()
```

At the end of the computation, the directory `output_dir` will contain 
simulated data usable by RACES. 
