---
title: "Tissue sampling"
output: rmarkdown::html_vignette
description: >
  How to sample cells from a simulated tissue
vignette: >
  %\VignetteIndexEntry{Tissue sampling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> *Disclaimer:* RACES/rRACES internally implements the probability
> distributions using the C++11 random number distribution classes whose
> algorithms are not defined by the standard. Thus, the simulation
> depends on the compiler used to compile rRACES, and because of that,
> the results reported in this article may differ from those obtained by
> the reader.

Once one has familiarised on how a tumour evolution simulation can be
programmed using rRACES (see `vignette("tissue_simulation")`), the next
step is to augment the simulation with sampling of tumour cells. This
mimics a realistic experimental design where we gather tumour sequencing
data.

This vignette introduces sampling using different type of models;
starting from simpler up to more complex simulation scenarios we
consider:

-   *multi-region sampling*: where at every time point multiple
    spatially-separated samples are collected;

-   *longitudinal sampling*: where the sampling is repeated at multiple
    time-points.

## Custom multi-region sampling

We consider a simple monoclonal model, without epimutants.

```{r, eval=TRUE}
library(rRACES)

# set the seed of the random number generator
set.seed(0)

# Monoclonal model, no epimutants
sim <- Simulation("Monoclonal")

sim$add_mutant(name = "A", growth_rates = 0.1, death_rates = 0.01)

sim$place_cell("A", 500, 500)
sim$run_up_to_size("A", 60000)
```

```{r,  fig.height=4, fig.width=3}
current <- plot_tissue(sim)
current
```

A sample is defined by a name and a bounding box, which has a $(x,y)$
coordinate for the bottom left point, and for the top right point.

For this simulation, we define two samples with names `"S_1_2"` and
`"S_1_2"`.

```{r,  fig.height=4, fig.width=3}
# We collect a squared box of (bbox_width x bbox_width) cells
bbox_width <- 50

# Box A1
bbox1_p <- c(400, 400)
bbox1_q <- bbox1_p + bbox_width

# Box B1
bbox2_p <- c(500, 500)
bbox2_q <- bbox2_p + bbox_width

library(ggplot2)

# View the boxes
current +
  geom_rect(xmin = bbox1_p[1], xmax = bbox1_q[2],
            ymin = bbox1_p[1], ymax = bbox1_q[2],
            fill = NA, color = "black") +
  geom_rect(xmin = bbox2_p[1], xmax = bbox2_q[2],
            ymin = bbox2_p[1], ymax = bbox2_q[2],
            fill = NA, color = "black")

# Sampling
sim$sample_cells("S_1_1", bottom_left = bbox1_p, top_right = bbox1_q)
sim$sample_cells("S_1_2", bottom_left = bbox2_p, top_right = bbox2_q)
```

> *Note:* Sampling removes cells from the tissue, as if the tissue was
> surgically resected. Therefore, cells that are mapped to the bounding
> box after application of `Simulation$sample_cells()` are no longer
> part of the simulation.

A new call to `plot_tissue()` will show the box where the cells have
been removed to be white.

```{r,  fig.height=4, fig.width=3}
plot_tissue(sim)
```

This is also reflected by `Simulation$get_cells()`, which now will not
find any tumour cell in the sampled region.

```{r}
library(dplyr)

# This should be empty
sim$get_cells(c(400, 400), c(400 + bbox_width, 400 + bbox_width)) %>% head
```

It should also be remarked that the sampling process exclusively
collects tumour cells, while it excludes wild-type cells.

### Cell division tree for sampled cells

Every sampled cell is linked, at the evolutionary level, to the other
cells that originate from the same initial cell. It helps to visualise
the evolutionary information on the cells that we have sampled as a
forest of trees (if one seeded multiple initial cells). The forest is an
object of the S4 class `SamplesForest`.

```{r}
forest <- sim$get_samples_forest()

forest
```

The forest has methods to obtain the nodes of the sampled cells.

```{r}
forest$get_nodes() %>% head
```

The leaves of the forest are sampled cells, while the internal nodes are
their ancestors. The field `sample` is not available for internal nodes,
and reports the sample name otherwise.

```{r}
# The leaves in the forest represent sampled cells
forest$get_nodes() %>%
  filter(!is.na(.data$sample)) %>%
  head
```

The roots of the forest having no ancestors.

```{r}
# If it is one cell, than the forest is a tree
forest$get_nodes() %>%
  filter(is.na(.data$ancestor))
```

We can also query the forest about the samples used to build it.

```{r}
forest$get_samples_info()
```

We can visualise the forest. This plot reports the cells and, on the
y-axis, their time of birth.

```{r,  fig.height=6, fig.width=11}
plot_forest(forest)
```

The plot shows also samples annotations and species but, for a large
number of cells, it might be complicated to view the full tree, unless a
very large canvas is used. For this reaason, it is possible to subset
the tree.

```{r,  fig.height=6, fig.width=11}
# Extract the subforest linked to sample
S_1_1_forest <- forest$get_subforest_for("S_1_1")

plot_forest(S_1_1_forest)
```

In general, these plots can be annotated with extra information, such as
the sampling times, and the MRCAs of each sample in the tree.

```{r,  fig.height=6, fig.width=11}
# Full plot
plot_forest(forest) %>%
  annotate_forest(forest)
# S_1_1 plot
plot_forest(S_1_1_forest) %>%
  annotate_forest(S_1_1_forest)
```

## Randomised multi-region samples

```{r}
# set the seed of the random number generator
set.seed(0)

sim <- Simulation("Randomised")

sim$add_mutant(name = "A", growth_rates = 0.1, death_rates = 0.01)

sim$place_cell("A", 500, 500)
sim$run_up_to_size("A", 60000)
```

We include a new mutant and let it grow. This new mutant has much higher
growth rates than its ancestor.

```{r,  fig.height=4, fig.width=3}
# Add a new mutant
sim$add_mutant(name = "B", growth_rates = 1, death_rates = 0.01)
sim$mutate_progeny(sim$choose_cell_in("A"), "B")

sim$run_up_to_size("B", 10000)

current <- plot_tissue(sim)
current
```

Since mutant start has been randomised by `Simulation$choose_cell_in()`,
we have no exact idea of where to sample to obtain for example, $100$ of
its cells. We can look visually at the simulation, but this is slow.

rRACES provides a `Simulation$search_sample()` function to sample
bounding boxes that contain a desired number of cells. The function
takes in input:

-   a bounding box size;
-   the number $n$ of cells to sample for a species of interest.

`Simulation$search_sample()` will attempt a fixed number of times to
sample the box, starting from positions occupied by the species of
interest. If a box that contains at least $n$ cells is not found within
a number of attempts, then the one with the largest number of samples is
returned.

This allows to program sampling without having a clear idea of the
tissue conformation.

```{r,  fig.height=4, fig.width=3}
# A bounding box 50x50 with at least 100 cells of species B
n_w <- n_h <- 50
ncells <- 0.8 * n_w * n_h

# Sampling ncells with random box sampling of boxes of size n_w x n_h
bbox <- sim$search_sample(c("B" = ncells), n_w, n_h)

# plot the bounding box
current +
  geom_rect(xmin = bbox$lower_corner[1], xmax = bbox$upper_corner[1],
            ymin = bbox$lower_corner[2], ymax = bbox$upper_corner[2],
            fill = NA, color = "black")
```

```{r}
# sample the tissue
sim$sample_cells("S_2_1", bbox$lower_corner, bbox$upper_corner)
```

Something similar with species `A`.

```{r,  fig.height=4, fig.width=3}
bbox <- sim$search_sample(c("A" = ncells), n_w, n_h)

# plot the bounding box
current +
  geom_rect(xmin = bbox$lower_corner[1], xmax = bbox$upper_corner[1],
            ymin = bbox$lower_corner[2], ymax = bbox$upper_corner[2],
            fill = NA, color = "black")
```

```{r}
# sample the tissue
sim$sample_cells("S_2_2", bbox$lower_corner, bbox$upper_corner)
```

The two samples have been extracted.

```{r,  fig.height=4, fig.width=3}
plot_tissue(sim)
```

Cell division tree.

```{r,  fig.height=6, fig.width=11}
forest <- sim$get_samples_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

## Randomised cell sampling (Liquid biopsy)

rRACES supports randomized cell sampling over the full tissue or a
rectangle thereof.

```{r}
# collect up to 2500 tumour cells randomly selected over the whole tissue
sim$sample_cells("S_2_3", num_of_cells = 2500)

bbox <- sim$search_sample(c("A" = ncells), n_w, n_h)

# collect up to 200 tumour cells randomly selected in the provided
# bounding box
sim$sample_cells("S_2_4", bbox$lower_corner, bbox$upper_corner, 200)

forest <- sim$get_samples_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

## Two populations with epigenetic state

We are now ready to simulate a model with epigenetic switches and
subclonal expansions.

```{r,  fig.height=4, fig.width=3}
# set the seed of the random number generator
set.seed(0)

sim <- Simulation("Two Populations")

sim$death_activation_level <- 20

# First mutant
sim$add_mutant(name = "A",
               epigenetic_rates = c("+-" = 0.01, "-+" = 0.01),
               growth_rates = c("+" = 0.1, "-" = 0.08),
               death_rates = c("+" = 0.1, "-" = 0.01))

sim$place_cell("A+", 500, 500)
sim$run_up_to_size("A+", 1000)
plot_tissue(sim, num_of_bins = 500)
```

We sample before introducing a new mutant.

```{r,  fig.height=4, fig.width=3}
bbox_width <- 10

sim$sample_cells("S_1_1",
                 bottom_left = c(480, 480),
                 top_right = c(480 + bbox_width, 480 + bbox_width))

sim$sample_cells("S_1_2",
                 bottom_left = c(500, 500),
                 top_right = c(500 + bbox_width, 500 + bbox_width))

plot_tissue(sim, num_of_bins = 500)
```

```{r,  fig.height=4, fig.width=3}
# Let it grow a bit more
sim$run_up_to_time(sim$get_clock() + 15)

plot_tissue(sim, num_of_bins = 500)
```

Add a new submutant.

```{r,  fig.height=4, fig.width=3}
cell <- sim$choose_cell_in("A")

sim$add_mutant(name = "B",
               epigenetic_rates = c("+-" = 0.05, "-+" = 0.1),
               growth_rates = c("+" = 0.8, "-" = 0.3),
               death_rates = c("+" = 0.05, "-" = 0.05))

sim$mutate_progeny(cell, "B")

# let it grow more time units
sim$run_up_to_size("B+", 7000)

plot_tissue(sim, num_of_bins = 500)
```

Sample again and plot the tissue

```{r,  fig.height=4, fig.width=3}
n_w <- n_h <- 25
ncells <- 0.9 * n_w * n_h

bbox <- sim$search_sample(c("A" = ncells), n_w, n_h)
sim$sample_cells("S_2_1", bbox$lower_corner, bbox$upper_corner)

bbox <- sim$search_sample(c("B" = ncells), n_w, n_h)
sim$sample_cells("S_2_2", bbox$lower_corner, bbox$upper_corner)

plot_tissue(sim, num_of_bins = 500)
plot_muller(sim)
```

Now we show the cell division tree, which starts being rather
complicated

```{r,  fig.height=12, fig.width=17}
forest <- sim$get_samples_forest()

plot_forest(forest) %>%
  annotate_forest(forest)
```

### Storing Samples Forests

A samples forest can be saved in a file by using the method
`SamplesForest$save()`.

```{r, echo=FALSE,  results='hide', message=FALSE}
unlink("samples_forest.sff")
```

```{r}
# check the file existence. It should not exists.
file.exists("samples_forest.sff")

# save the samples forest in the file "samples_forest.sff"
forest$save("samples_forest.sff")

# check the file existence. It now exists.
file.exists("samples_forest.sff")
```

The saved samples forest can successively be load by using the function
`load_samples_forest()`.

```{r}
# load the samples forest from "samples_forest.sff" and store it in `forest2`
forest2 <- load_samples_forest("samples_forest.sff")

# let us now compare the samples forests stored in `forest` and `forest2`;
# they should be the same.
forest
forest2
```
