---
title: "Sequencing Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sequencing Simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE, results='hide', message=FALSE}
unlink("rRACES_SAM", recursive = TRUE)
```

Once a phylogenetic forest has been computed (see `vignette("mutations")`)
for the details), rRACES/RACES can simulate the sequencing of the 
samples in the forest and return the observed data.

### Sequencing Simulation

Let us consider the phylogenetic forest produced in 
`vignette("mutations")`.

```{r}
library(rRACES)

phylo_forest <- load_phylogenetic_forest("phylo_forest.sff")

phylo_forest
```

The loaded phylogenetic forest models the cell evolution of 
4 different samples: `S_1_1`, `S_1_2`, `S_2_1`, and `S_2_2`. 

We can simulate the sequencing of these samples with coverage 
of 2.5X by as it follows. 

```{r}
# let us simulate a 2.5x sequencing of the four sample
seq_results <- simulate_seq(phylo_forest, coverage = 2.5)

# let us load the dplyr library to filter the `simulate_seq` output
library(dplyr)

seq_results %>% head
```

The function `simulate_seq` returns a data frame whose rows 
represent the mutations observed in the simulated reads
(currently, only SNVs are supported).

The number of columns in the data frame depends on the number 
of samples in the phygenetic forest. The first 4 columns 
describe the mutation and report the chromosome and the position 
in the chromosome of the mutation (columns "`chromosome`" and 
"`chr_pos`", respectively), the genetic context in which the 
mutation occurs (column "`context`"), and the new base (column 
"`mutated_base`").
Then, there are 3 columns for each of the samples: they 
contain the number of simulated reads affected by the 
mutation, the sequencing coverage of the mutation locus, 
and the ratio between these two last columns 
(columns "`<sample name> occurrences`", "`<sample name> coverage`", 
and "`<sample name> VAF`", respectively).

### Cell Partition by Epigenetic State (Epi-FACS)

The sequencing simulation supports cell partition by epigenetic state (epi-FACS) by 
using the `simualate_seq` parameter `epi_FACS`.
At the beginning of the sequencing simulation process, each sample in the phylogenetic 
forest is splitted in two distinct samples: `<sample name>_P`, that contains 
all the cells in the original sample having epigenetic state `+`, and 
`<sample name>_N`, that consists in the cells having epigenetic state `-`.
The results of the simulation reports the data for the splitted samples.

```{r}
seq_results <- simulate_seq(phylo_forest, coverage = 2.5, epi_FACS = TRUE)

seq_results %>% head
```


### Saving the Simulated Reads

rRACES/RACE can also save the simulated reads in the SAM format 
(see [@10.1093/bioinformatics/btp352]). By setting the 
optional parameter `write_SAM` to `TRUE`, 
the function `simulate_seq` creates the directory `rRACES_SAM`
and saves the SAM files in it. Each file is named after one the 
reference genome chromosomes and it contains the reads simulated 
for it. 

```{r}
seq_results <- simulate_seq(phylo_forest, coverage = 2.5, write_SAM = TRUE)

SAM_files <- list.files("rRACES_SAM/")

ex_file <- paste("rRACES_SAM/", SAM_files[1], sep="")
for (line in readLines(ex_file, n = 10)) {
    cat(paste(line,"\n"))
}
```

Each SAM file contains the reads produced by simulating the sequencing 
of *all* the samples. The command line tools `samtools` can be used to
split the reads by sample (see [here](http://www.htslib.org)).

```{bash, eval = FALSE}
foo@bar % samtools split  -f "%*_%\!.sam" rRACES_SAM/chr_22.sam
foo@bar % ls chr_22_*
chr_22_S_1_1.sam        chr_22_S_1_2.sam        chr_22_S_2_1.sam        chr_22_S_2_2.sam
```

The resulting files are named after the phylogenetic forest samples and each of them 
contains the reads of only one sample.

Please, refer to the [`samtools split` manual](http://www.htslib.org/doc/samtools-split.html)
for more details.


```{r, echo=FALSE, results='hide', message=FALSE}
unlink("rRACES_SAM", recursive = TRUE)
unlink("demo", recursive = TRUE)
unlink("phylo_forest.sff")
```
