---
title: "Time Series Analysis"
output: rmarkdown::html_vignette
description: >
  Analyse the evolution of the system over time
vignette: >
  %\VignetteIndexEntry{Time Series Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(rRACES)

sim <- new(Simulation, "Test")

sim$death_activation_level <- 50

# Set the tissue name and size to different values
sim$update_tissue("Liver", 1200, 900)

sim$add_mutant(name = "A",
               epigenetic_rates = c("+-" = 0.01, "-+" = 0.01),
               growth_rates = c("+" = 0.2, "-" = 0.08),
               death_rates = c("+" = 0.1, "-" = 0.01))

# We add one cell of species A+ (mutant A in epistate +) in
# position (500, 500).
sim$place_cell("A+", 500, 500)

sim$run_up_to_size("A+", 500)

sim$run_up_to_event("death", "A-", 100)

# Run the simulation for other 15 time units
sim$run_up_to_time(sim$get_clock() + 15)

# Add mutant
sim$add_mutant(name = "B",
               epigenetic_rates = c("+-" = 0.05, "-+" = 0.05),
               growth_rates = c("+" = 0.3, "-" = 0.3),
               death_rates = c("+" = 0.05, "-" = 0.1))

# Mutant injection
sim$mutate_progeny(sim$choose_cell_in("A"), "B")

# New evolution
sim$run_up_to_size("B+", 600)

# Define evolutionary parameters
sim$add_mutant(name = "C",
               epigenetic_rates = c("+-" = 0.1, "-+" = 0.1),
               growth_rates = c("+" = 0.2, "-" = 0.4),
               death_rates = c("+" = 0.1, "-" = 0.01))

# Choose and mutate
sim$mutate_progeny(sim$choose_cell_in("A",  c(450, 550), c(500, 600)), "C")

sim$run_up_to_time(sim$get_clock() + 7)
```

> *Note:* This article presents advanced topics on tissue simulation.
> Refer to `vignette("tissue_simulation")` for an introduction on the subject.

> *Disclaimer:* RACES/rRACES internally implements the
> probability distributions using the C++11 random number
> distribution classes. The standard does not specify their
> algorithms, and the class implementations are left free
> for the compiler. Thus, the simulation output depends on
> the compiler used to compile RACES, and because of that,
> the results reported in this article may differ from
> those obtained by the reader.

Sometimes it is convenient to plot a time series of a simulation,
reporting species or firing counts over time. Since rRACES is
programmable, it is immediate to make a for-loop algorithm and collect
the simulation data over time.

## Default History-Based Data

However, this is not required because at the end of any `run_to_*`
methods, RACES stores the data about the number of species cells, and
that of event firings. These data can be extracted.

Let us consider the simulation `sim` as produced in
`vignette("tissue_simulation")`.

```{r}
library(rRACES)
library(dplyr)

# The firings
sim$get_firing_history() %>% head()

# For example, total number of the deaths on `B+` at the end of the
# previous calls of the `run_to_*` methods
sim$get_firing_history() %>%
  filter(event == "death", mutant == "B", epistate == "-")

# The counts
sim$get_count_history() %>% head()
```

The time-series can be plot using `plot_timeseries()`

```{r,  fig.height=4, fig.width=4}
# Time-series plot
plot_timeseries(sim)
```

## Custom Time-Series

If the default time-series is not enough coarse-grained, one can set
`Simulation$history_delta` to increase the sampling rate of
the state (by default, `Simulation$history_delta` is set to $0$).

We show this by re-simulating a tumour with two submutants.

```{r}
# Example time-series on a new simulation, with coarse-grained time-series
sim <- new(Simulation, "Finer Time Series")

sim$history_delta <- 1
sim$death_activation_level <- 100

# A
sim$add_mutant(name = "A",
               epigenetic_rates = c("+-" = 0.01, "-+" = 0.01),
               growth_rates = c("+" = 0.2, "-" = 0.08),
               death_rates = c("+" = 0.1, "-" = 0.01))

sim$place_cell("A+", 500, 500)
sim$run_up_to_size("A+", 400)

# B (linear inside A)
sim$add_mutant(name = "B",
               epigenetic_rates = c("+-" = 0.05, "-+" = 0.05),
               growth_rates = c("+" = 0.3, "-" = 0.3),
               death_rates = c("+" = 0.05, "-" = 0.1))

sim$mutate_progeny(sim$choose_cell_in("A"), "B")

sim$run_up_to_size("B-", 300)

# C (linear inside B)
sim$add_mutant(name = "C",
               epigenetic_rates = c("+-" = 0.1, "-+" = 0.1),
               growth_rates = c("+" = 0.2, "-" = 0.4),
               death_rates = c("+" = 0.1, "-" = 0.01))

sim$mutate_progeny(sim$choose_cell_in("B"), "C")

# D (linear inside A, so branching with C) - same parameters of C
sim$add_mutant(name = "D",
               epigenetic_rates = c("+-" = 0.1, "-+" = 0.1),
               growth_rates = c("+" = 0.2, "-" = 0.4),
               death_rates = c("+" = 0.1, "-" = 0.01))

sim$mutate_progeny(sim$choose_cell_in("A"), "D")

sim$run_up_to_size("D+", 1000)
```

The time-series can be plot using `plot_timeseries()`.

```{r,  fig.height=4, fig.width=4}
# Time-series plot
plot_timeseries(sim)

# Logscale helps seeing the different effective growth rates
plot_timeseries(sim) + ggplot2::scale_y_log10()
```

### Muller Plot

We can also get a Muller plot of the evolution using
[ggmuller](https://CRAN.R-project.org/package=ggmuller).

```{r,  fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
# default Muller plot
plot_muller(sim)
```

In this case every population is annotated as a descendant of the
ancestor mutant. Note however that reversible espistates do not fit a
traditional Muller plot because they violate the no-back mutation model.

In this case, _rRACES_ will show first the epistate that was randomly
injected in the simulation, and the second will result by linear. This
is not a completely correct perspective of the simulation time-series;
still, it help understand trends.

```{r,  fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
# Custom Mullers
clock <- sim$get_clock()

plot_muller(sim) + ggplot2::xlim(clock * 3/4, clock)

plot_muller(sim) +
  ggplot2::xlim(clock * 3/4, clock) +
  ggplot2::scale_y_log10()
```

## Time-Varying Evolutionary Rates

You can model the fact that the rates of one species. For instance, this
happens when a population is subject to a targeted treatment.

Considering the example above, where `C` and `D` have the same rates, we
increase the death rate of both `C+` and `C-` species, as well as `B+`
and `B-`.

```{r,  fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
# Current rates
sim

# Raise the death rate levels
sim$update_rates("B+", c(death = 3))
sim$update_rates("B-", c(death = 3))
sim$update_rates("C+", c(death = 3))
sim$update_rates("C-", c(death = 3))

# Now D will become larger
sim$run_up_to_size("D+", 6000)

# Current state
sim

# This now show the change in rates
clock <- sim$get_clock()
plot_muller(sim) + ggplot2::xlim(clock * 3/4, clock)

plot_muller(sim) +
  ggplot2::xlim(clock * 3/4, clock) +
  ggplot2::scale_y_log10()
```

